Unimodular covers of 3-dimensional parallelepipeds and Cayley sums

### 2019-05-26 unimodular-covers.tex
typos
finished rough draft of section 2

\newcommand{\bb}{{\mathbf{b}}}
\newcommand{\cc}{{\mathbf{c}}}
\newcommand{\ttt}{\mathbf{t}}
\newcommand{\GL}{{GL}_r (K)}
\newcommand{\Gin}{\ensuremath{\mathrm{Gin}}}
\newcommand{\gin}{\ensuremath{\mathrm{gin}}}
\newcommand{\GIN}{\ensuremath{\mathrm{GIN}}}
\newcommand{\Gins}{\ensuremath{\mathrm{Gins}}}
\newcommand{\gins}{\ensuremath{\mathrm{gins}}}
\newcommand{\Lex}{\ensuremath{\mathrm{Lex}}}
\newcommand{\dele}[1]{\ensuremath{\Delta^e({#1})}}
\newcommand{\dels}[1]{\ensuremath{\Delta^s({#1})}}
\newcommand{\delc}[1]{\ensuremath{\Delta^c({#1})}}
\newcommand{\Shift}{\ensuremath{\mathrm{Shift}}}

\renewcommand{\int}{\operatorname{int}}
\newcommand{\width}{\operatorname{width}}
\newcommand{\length}{\operatorname{length}}

\newcommand{\cone}{\ensuremath{\mathrm{cone}}\hspace{1pt}}
\newcommand{\conv}{\ensuremath{\mathrm{conv}}\hspace{1pt}}
\newcommand{\cayley}{\operatorname{Cay}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\reg}{\operatorname{reg}}
\newcommand{\depth}{\operatorname{depth}}

@@

\begin{definition}
Let $T$ be a tetrahedron with vertex set $p_1,p_2,p_3,p_4$. Consider points $q_1,q_2,q_3,q_4$ defined as $q_i= \frac12 (p_1+p_2+p_3+p_4) - p_i$ and define $C(T)=\conv(p_i,q_i: i\in[4])$. $C(T)$ is a parallelepiped whose faces are of the form $\conv(p_i, p_j, q_k, q_l)$ for all $\{i,j,k,l\}=[4]$. We call $C(T)$ the \emph{parallelepiped circumscribed} to $T$.

@@

Some properties of this construction:

1) $S$ is ``half a fundamental domain'' for the affine lattice $\Lambda_P$ generated by $p_1,\dots,p_4$, in the following precise sense: if $x$ is an arbitrary point, then $S$ intersects either $\Lambda_P + x$ or $\Lambda_P -x$ (and, if $x$ is generic then the intersection is unique). Proof: think of this property for the standard cube and the lattice $\{sum of coordinates is even\}$ generated by the vertices of the regular tetrahedron inscribed in it. (Remark: let us assume without loss of generality that $\Lambda_P$ contains the origin, so that $\Lambda + x$ is the class of $x$ modulo $\Lambda$).

@@

\begin{proof}
By White's classification of empty tetrahedra (\cite{White1964}, see also, e.~g.~\cite[Sect.~4.1]{HPPS-survey}), there is no loss of generality in assuming $T=\conv(p_1,p_2,p_3,p_4)$ with
\[
p_1=(0,0,0), \quad
p_2=(1,0,0), \quad
p_3=(0,0,1), \quad
p_4=(a,b,1).
\]
where $b\ge 2$ is the (normalized) volume of $T$, and $a\in \{1,\dots,b-1\}$ satisfies $\gcd(a,b)=1$. This gives 
\paco{picture of the quadrilaterals $p_1q_4p_2q_3$ and $q_2p_3q_1p_4$ would be good.}
\begin{align*}
q_1=\left(\frac{1+a}2,\frac{b}2,1\right), &&
q_2=\left(\frac{a-1}2,\frac{b}2,1\right), \quad\\
q_3=\left(\frac{1+a}2,\frac{b}2,0\right), &&
q_4=\left(\frac{a-1}2,-\frac{b}2,0\right).
\end{align*}

@@

\begin{proof}[Proof of \Cref{thm:parallelepipeds}]
Let $P \subseteq \R^3$ be a parallelopiped. It can be triangulated into empty lattice tetrahedra, simply by taking a triangulation which uses all lattice points as vertices.  If all the tetrahedra in the triangulation are unimodular we are done. Otherwise, suppose $T$ is be empty tetrahedron in the triangulation which is not unimodular. \Cref{lemma:3<4} guarantees that one of the corner tetrahedra $T_i$ of the parallelepiped $C(T)$ circumscribed to $T$ is contained in $P$. Without loss of generality, suppose $T_4 = \conv(p_1, p_2, p_3,q_4)$ is in $P$. By \Cref{lemma:corner}, we know that $T_4$ contains a lattice point, which we denote by $l$.

Then $\conv(T\cup \{p_5\})$ can be triangulated into 
\[
\conv(p_1,p_2,p_4, l),
\conv(p_1,p_3,p_4, l),
\conv(p_2,p_3,p_4, l).
\]

Since these all have volume strictly smaller than $T$ \giulia{not convinced this is so obvious or true for any $l$: the volume of $\conv(T \cup q_4)$ is 3/2 times that of $T$, and if two of the three have very small volume... But I think (perhaps check in coordinates) we can choose $l$ wisely.}, this allows us to cover $T$ with lattice simplices of smaller volume recursively, until it is covered by unimodular ones.
\end{proof}

@@

In turn, our proof of \Cref{lemma:3<4} relies on the fact that parallelepipeds have only three (pairs of) normal vectors. The proof fails if $P$ has four of them, as the following non-IDP octahedron and triangular prism show:
\[
\text{ write the octa and tri prism}.
\]
In both cases, the only lattice points are the six vertices and the origin. The point $(1,1,1)$ lies in the second dilation but is not the sum of two lattice points in the polytope.
\end{remark}

>>>>

\newcommand{\bb}{{\mathbf{b}}}
\newcommand{\cc}{{\mathbf{c}}}
\newcommand{\ttt}{\mathbf{t}}
\newcommand{\GL}{{GL}_r (K)}

\renewcommand{\int}{\operatorname{int}}
\newcommand{\width}{\operatorname{width}}
\newcommand{\Vol}{\operatorname{Vol}}
\newcommand{\length}{\operatorname{length}}

\newcommand{\cone}{\ensuremath{\mathrm{cone}}\hspace{1pt}}
\newcommand{\conv}{\ensuremath{\mathrm{conv}}\hspace{1pt}}
\newcommand{\cayley}{\operatorname{Cay}}
\newcommand{\rank}{\operatorname{rank}}
\newcommand{\reg}{\operatorname{reg}}

@@

\begin{definition}
Let $T$ be a tetrahedron with vertex set $p_1,p_2,p_3,p_4$. Consider points $q_1,q_2,q_3,q_4$ defined as $q_i= \frac12 (p_1+p_2+p_3+p_4) - p_i$ and define

\[
C(T)=\conv(p_i,q_i: i\in[4]).
\]

$C(T)$ is a parallelepiped whose faces are of the form $\conv(p_i, p_j, q_k, q_l)$ for all $\{i,j,k,l\}=[4]$, which we call the \emph{parallelepiped circumscribed} to $T$.

@@

\Giulia{not sure if we need this remark:
$S$ is ``half a fundamental domain'' for the affine lattice $\Lambda_P$ generated by $p_1,\dots,p_4$, in the following precise sense: if $x$ is an arbitrary point, then $S$ intersects either $\Lambda_P + x$ or $\Lambda_P -x$ (and, if $x$ is generic then the intersection is unique). Proof: think of this property for the standard cube and the lattice $\{sum of coordinates is even\}$ generated by the vertices of the regular tetrahedron inscribed in it. (Remark: let us assume without loss of generality that $\Lambda_P$ contains the origin, so that $\Lambda + x$ is the class of $x$ modulo $\Lambda$).
}

@@

\begin{proof}
By White's classification of empty tetrahedra (\cite{White1964}, see also, e.~g.~\cite[Sect.~4.1]{HPPS-survey}), there is no loss of generality in assuming $T=\conv(p_1,p_2,p_3,p_4)$ with
\[
p_1=(0,0,0), \quad
p_2=(1,0,0), \quad
p_3=(0,0,1), \quad
p_4=(a,b,1).
\]
where $b\ge 2$ is the (normalized) volume of $T$, and $a\in \{1,\dots,b-1\}$ satisfies $\gcd(a,b)=1$. This gives 
\paco{picture of the quadrilaterals $p_1q_4p_2q_3$ and $q_2p_3q_1p_4$ would be good.}
\begin{align*}
q_1=\left(\frac{1+a}2,\frac{b}2,1\right), &&
q_2=\left(\frac{a-1}2,\frac{b}2,1\right), \quad\\
q_3=\left(\frac{1+a}2,\frac{b}2,0\right), &&
q_4=\left(\frac{1-a}2,-\frac{b}2,0\right).
\end{align*}

@@

\begin{proof}[Proof of \Cref{thm:parallelepipeds}]
Let $P \subseteq \R^3$ be a parallelopiped. It can be triangulated into empty lattice tetrahedra, simply by taking a triangulation which uses all lattice points as vertices.  If all the tetrahedra in the triangulation are unimodular we are done. Otherwise, suppose $T$ is be empty tetrahedron in the triangulation which is not unimodular. \Cref{lemma:3<4} guarantees that one of the corner tetrahedra $T_i$ of the parallelepiped $C(T)$ circumscribed to $T$ is contained in $P$. Without loss of generality, suppose $T_4 = \conv(p_1, p_2, p_3,q_4)$ is in $P$. By \Cref{lemma:corner}, we know that $T_4$ contains a lattice point, which we denote by $l$. Observe that from the proof of \Cref{lemma:corner} we know more: $l$ lies in the face $\conv(p_1, p_2, q_4)$.\giulia{we need this for the volume computation below. perhaps it should be made explicit in the statemente of the lemma...}

Then $S=\conv(T\cup \{l\})$ can be triangulated in two different ways: $S=T \cup T'_4 = S_1 \cup S_2 \cup S_3$, where $T'_4 = \conv(p_1, p_2, p_3, l) \subseteq T_4$ and
\[
S_1= \conv(p_2,p_3,p_4, l),
S_2=\conv(p_1,p_3,p_4, l),
S_3=\conv(p_1,p_2,p_4, l).
\]

Each of the tetrahedra $S_i$ has lattice volume strictly smaller than $T$: indeed, $\Vol(S_3)=\Vol(T'_4)$, since these tetrahedra have a common face $\conv(p_1, p_2, l)$ and the remaining vertex is at distance one from this face in both. Now $T'_4$ is contained in $T_4$, and we know that each corner tetrahedron $T_i$ has volume half that of $T$. The remaining two tetrahedra $S_1$ and $S_2$ must therefore have volume summing to that of $T$, and therefore each has volume strictly less than $T$.

This allows us to cover $T$ with lattice simplices of smaller volume recursively, until it is covered by unimodular ones.
\end{proof}

@@

In turn, our proof of \Cref{lemma:3<4} relies on the fact that parallelepipeds have only three (pairs of) normal vectors. The proof fails if $P$ has four of them, as the following non-IDP octahedron and triangular prism show:

\begin{example}
\label{ex:non-IDP}
The following lattice octahedron is not IDP:
\[
Q= \conv((0,1,1),(1,0,1),(1,1,0),(0,-1,-1),(-1,0,-1),(-1,-1,0))
\]

The following lattice triangular prism is not IDP:
\[
P=\conv((0,1,1),(1,0,1),(1,1,0),(-1,0,0),(0,-1,0),(0,0,-1)).
\]
Indeed, in both polytopes the only lattice points are the six vertices and the origin. The point $(1,1,1)$ lies in the second dilation but is not the sum of two lattice points in the polytope.
\giulia{is it necessary to be more explicit, for example show that these are the only lattice points?}
\end{example}

\end{remark}
